# Data sources
---
title: "Data Sources"
author: "Rahul Subramaniam, Santos Hernandez"
output:
  html_document: default
  pdf_document: default
---

```{r, include=TRUE}
SOLUTIONS <- FALSE
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = SOLUTIONS, 
                      eval = SOLUTIONS, cache = FALSE)
library(tidyverse)
```



```{r}
# keep this chunk in your .Rmd file
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


The NBA tools are available as a part of the nbaStatR package. We install it using the devtools package offered by github.


```{r}
library(nbastatR)
library(future)
plan(multiprocess) 
```



c) Sampole view of Data 
```{r}
game_logs(seasons = 2010:2019)
```




## Game statistics
In the NBA, the largest source of game statistics comes from the league itself. This data is 
publicly available at the [NBA website](https://www.nba.com/stats/), with some stats going back as far as 1946.
Scorekeepers are hired to watch every single game, tracking every statistic that is available to us. Many of the 
statistics are raw counts of things like points, minutes played, and rebounds, while others are 
calculated metrics. This does not discount the fact that the stats we see are 
subject to [human error](https://deadspin.com/the-confessions-of-an-nba-scorekeeper-5345287),
but generally, the NBA is the best data source available for game statistics. 

[NBAstatR](https://github.com/abresler/nbastatR) is a package that crawls the website to return the data directly in R. The game data is aggregated 
and rolled up based on specified parameters. For example, one could request the per-game averages 
of each team for the 2015 season, from March to April. The package would return a table 
with a row for each team, with average statistics for allgames played in that time period. This reflects the behavior of the NBA stats 
website, which does the same thing. There are 30 teams in the modern NBA, so 30 rows are always returned in requests for team statistics. 
When requesting player data, the number of rows can vary because not every player plays every game and people are constantly getting added/dropped from rosters. 
Generally, for the whole regular season, there are between 500-550 rows of unique player data. Less are returned for 
requests of smaller windows of time, as some players may not play at all during that time. 

Basketball stats are complicated. There are a seemingly endless amount of statistics 
that are available for each player and team. For example, the "Traditional" stats reported include things 
like 3 Point Field Goals Made, Free Throw Percentage, Rebounds, Assists, and Steals, among many others. 
For our project, we were most interested understanding _general_ team/player performance, which can be 
difficult to discern from all of those statistics. So, we focused on the 
Offensive, Defensive, and Net ratings. The Offensive rating can be understood as 
the amount of points produced on offense when the player was on the court, while Defensive rating can be understood as 
the amount of points allowed on defense when the player was on the court. The way these are 
calculated are somewhat [complicated](https://www.basketball-reference.com/about/ratings.html), 
but are a good way to understand general performance. Team versions of these ratings are calculated 
based on the underlying player statistics.

### COVID cases

A single table was scraped from a [basketball magazine's website](https://www.slamonline.com/nba/nba-coronavirus-covid-list/) 
(SLAM) to get a list of players that were confirmed to have COVID, and when it was reported by the magazine.

### Kobe Bryant's Connections 

A 1 Dimensional list of players was obtained from a [Celebrity lifestyle website](https://www.thethings.com/15-nba-stars-kobe-bryant-was-close-to/) (The Things) to get a list of players who were close friends of Kobe Bryant.

## Player positions and injuries

We also needed other data that went beyond game statistics: player position data and injury data. 
The NBA also makes this data somewhat available on its website, does not aggregate and release it 
as conveniently. Player position data is limited only to forward, guard, and center, 
while most think of the game as played in 5 positions - point guard, shooting guard, small forward, power forward, and center. 
Injury data available on the NBA website is only for current injuries, and is not historical. 

////EXPAND VVV
We turned to [Basketball Reference](https://www.basketball-reference.com/about/ratings.html) for more complete player profiles. T
his data is collected by a parent organization that manages 
several sports databases. The core of its historical data [comes from](https://www.basketball-reference.com/about/sources.html) 
Sean Lahman, a researcher/journalist famous for compiling open source sports data and encyclopedias.

Injury data was scraped from [Pro Sports Transactions](http://www.prosportstransactions.com/), which compiles data that is released by the NBA.

## Issues

When we compare a "month" of game data for each team, each team may not have played 
the same amount of games due to scheduling. The sample for each average can be 
slightly unequal by +/- a game played from team to team. The solution to this would be to obtain averages based 
on the number of games played before a certain date, but the API lacks this slicing capability. 
It might be possible pull the box scores of every single game, count out and filter based on date, and then create the averages manually, 
but this approach would be a significant amount of work for perhaps a small change in results. As previously stated, 
we are looking for differences in general performance.
Inconsistency was observed in the naming conventions of the basketball positions. E.g. the position scraped for a player was observed to be Guard, but on a study of that players position,it was found out that he played Point Guard.
Many players have a nicknames/unofficial names that a few of the data sources use.Hence, keeping track of player information while subsetting and filtering out data was an issue.

