# Data transformation

## Game statistics

### COVID cases

When using the NBA stats R package, requested data comes as a nested dataframe. 
Each row has data for a single season, with either regular season or playoff data. 
So if you requested 2018-19 and 2019-2020 data for the regular season and playoffs, 
4 rows would be returned. A column at the end has the actual game stats dataframes,
so these must be unnested.
```{r eval = FALSE, echo = TRUE}
covid_season_month_before <- teams_players_stats(seasons = 2020, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "2-11-20", date_to = "3-11-20") %>% 
  unnest(dataTable) %>% 
  mutate(df_type = "month_before")

bubble <- teams_players_stats(seasons = 2020, measures = "Advanced",  types = "team", season_types = c("Regular Season", "Playoffs"), tables = "general", modes = "PerGame", date_from = "7-29-20", date_to = "10-11-20") %>% 
  unnest(dataTable) 
```

We split the time periods that we wanted to compare into separate dataframes and added a new column, 
`df_type`, to distinguish them.
```{r eval = FALSE, echo = TRUE}
bubble_seeding <- bubble %>% 
  filter(typeSeason == "Regular Season") %>% 
  mutate(df_type = "bubble_seeding") 

bubble_playoffs <- bubble %>% 
  filter(typeSeason == "Playoffs") %>% 
  mutate(df_type = "bubble_playoffs")   
```

Finally, we made the data long for it to work with ggplot. The data when requested comes wide, 
with a column for each statistic.
```{r eval = FALSE, echo = TRUE}
bubble_stats_long <- bind_rows(covid_season_month_before,
                               bubble_seeding,
                               bubble_playoffs,) %>%
  select(nameTeam:paceE,
         df_type,
         -contains("rank")) %>% #drop some stats we're not interested in
  pivot_longer(gp:paceE, names_to = "rating", values_to = "value") %>%
  semi_join(bubble_teams, by = "nameTeam") #only teams that were invited to bubble
```

The same process was used for both team and player data because they come in the same form.


Some additional steps were taken to integrate the COVID case data. The table from SLAM was 
crawled with `rvest` and cleaned slightly to be able to join onto the player stats. Furthermore, 
the dates were converted into date objects so that they would display in order on the x-axis.
```{r eval = FALSE, echo = TRUE}
covid_players <- read_html("https://www.slamonline.com/nba/nba-coronavirus-covid-list/") %>% 
  html_element(css = ".wp-block-table") %>% 
  html_table(header = TRUE)

covid_players_fixed <- covid_players %>% 
  rename(name = "Player",
         date_reported = "Reported") %>% 
  mutate(team_abb = str_sub(name, -4, -2),
         name = str_remove(name, " \\(.+\\)"),
         date_reported = str_remove(date_reported, " \\(.+\\)"),
         date_reported = mdy(str_c(date_reported, " 2020")))

```

## Kobe Bryant Demise Effect

Similar to the format in the COVID cases problem, the data for every NBA team comes as a nested dataframe. However, this time the time frame of the data to be analyzed is much lesser (data ranges from 3 weeks before and 2 weeks after Kobe's death). Hence, we cannot request the data for all seasons in one go, and must selectively obtain the data for each season.

We obtained the data for every season in two separate dataframes: 

1. The data for every team before Kobe's demise ( range from 5th January to 25th of January)
2. The data for every team after Kobe's demise (range from 26th January to 9th of February)

For the sake of analysis and visualization, we had to divide the 30 NBA teams into top 15 and bottom 15 (based on the net rating score for every team in the 2020 season).

```{r eval = FALSE, echo = TRUE}
team_data_2020_after_26 <- teams_players_stats(seasons = 2020, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "1-26-20", date_to = "2-9-20")$dataTable[[1]]

team_data_2020_before_26 <- teams_players_stats(seasons = 2020, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "1-5-20", date_to = "1-25-20")$dataTable[[1]]

ordered_team_data_2020_after_death <-team_data_2020_after_26[order(team_data_2020_after_26[,41]),]


team_list <- ordered_team_data_2020_after_death$nameTeam
bottom_teams <- team_list[1:15]
top_teams <- team_list[16:30]
```

This was followed by a series of subsetting and pivoting operations so that we could get the net rating of teams within the pre bryant as well as the post bryant timeframe. We pivotted longer on the net rating variable


```{r eval = FALSE, echo = TRUE}
team_data_2020_after_26_top_teams <- subset(team_data_2020_after_26,nameTeam %in% top_teams)
netrating_after_26<- team_data_2020_after_26_top_teams$netrtg

teams <- team_data_2020_after_26_top_teams$nameTeam
bar_plot_df_2020_top_teams <- data.frame(teams,netrating_before_26,netrating_after_26)

pivoted_bar_plot_df_2020_top_teams <- pivot_longer(bar_plot_df_2020_top_teams,cols = c('netrating_before_26','netrating_after_26'),names_to='variable', 
values_to="value")

netrating_after_26<- team_data_2020_after_26_bottom_teams$netrtg
pivoted_bar_plot_df_2020_top_teams
```


A similar approach was taken to obtain team statistics for years before and after Kobe's death.I We retrieved the team performance statistics for previous years till 2017 season, and subsetted only those teams in our selected teams list.


```{r eval = FALSE, echo = TRUE}
selected_teams_years_analysis = c('Los Angeles Lakers','LA Clippers','Dallas Mavericks','Philadelphia 76ers','Utah Jazz')

team_data_2018_after_26 <- teams_players_stats(seasons = 2018, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "1-26-18", date_to = "2-9-18")$dataTable[[1]]
selected_team_data_2018_after_26 <- subset(team_data_2018_after_26,nameTeam %in% selected_teams_years_analysis) 

team_data_2017_after_26 <- teams_players_stats(seasons = 2017, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "1-26-17", date_to = "2-9-17")$dataTable[[1]]
selected_team_data_2017_after_26 <- subset(team_data_2017_after_26,nameTeam %in% selected_teams_years_analysis) 

team_data_2016_after_26 <- teams_players_stats(seasons = 2016, measures = "Advanced", types = "team", tables = "general", modes = "PerGame", date_from = "1-26-16", date_to = "2-9-16")$dataTable[[1]]
selected_team_data_2016_after_26 <- subset(team_data_2016_after_26,nameTeam %in% selected_teams_years_analysis) 

```
We then bind the row data for all the years together, and mutate the dataframe obtained with the corresponding year.

```{r eval = FALSE, echo = TRUE}

consolidated_selected_data <- rbind(selected_team_data_2020_after_26,selected_team_data_2019_after_26,selected_team_data_2018_after_26 ,selected_team_data_2017_after_26, selected_team_data_2016_after_26,selected_team_data_2021_after_26)

date<-as.POSIXct(consolidated_selected_data$rangeDateTo,format = '%m-%d-%Y')
year <- as.numeric(format(date,format = '%Y'))
consolidated_selected_data <- cbind(consolidated_selected_data,year)

```

Lastly, we obtain a list of Kobe's friends, and their gameplay statistics for the 2020 season and 2019 season (from 12th Jan to 16th February).

```{r eval = FALSE, echo = TRUE}
kobe_bryant_friends <- c('Carmelo Anthony','LeBron James','Anthony Davis','Trae Young','Kevin Durant','Stephen Curry','Chris Paul')

team_data_2020_after_26 <- teams_players_stats(seasons = 2020, measures = "Advanced", types = "player", tables = "general", modes = "PerGame", date_from = "1-26-20", date_to = "2-16-20")$dataTable[[1]]

team_data_2020_before_26 <- teams_players_stats(seasons = 2020, measures = "Advanced", types = "player", tables = "general", modes = "PerGame", date_from = "1-12-20", date_to = "1-25-20")$dataTable[[1]]

team_data_2019_after_26 <- teams_players_stats(seasons = 2019, measures = "Advanced", types = "player", tables = "general", modes = "PerGame", date_from = "1-26-19", date_to = "2-16-19")$dataTable[[1]]

team_data_2019_before_26 <- teams_players_stats(seasons = 2019, measures = "Advanced", types = "player", tables = "general", modes = "PerGame", date_from = "1-12-19", date_to = "1-25-19")$dataTable[[1]]
```

We then clean the data obtained by a series of subsetting operations, and then by lengthening the data using a row bind(). 

```{r}
selected_players_data_2020_after_26 <- subset(team_data_2020_after_26,namePlayer %in% kobe_bryant_friends)
selected_players_data_2020_before_26 <- subset(team_data_2020_before_26,namePlayer %in% kobe_bryant_friends)
selected_players_data_2019_after_26 <- subset(team_data_2019_after_26,namePlayer %in% kobe_bryant_friends)
selected_players_data_2019_before_26 <- subset(team_data_2019_before_26,namePlayer %in% kobe_bryant_friends)

consolidated_player_data <- rbind(selected_players_data_2020_after_26,selected_players_data_2020_before_26,selected_players_data_2019_after_26,selected_players_data_2019_before_26)

```

## Injury and Player positions

The data for the entire analysis is retrieved by iterating and scraping through 1,291 pages of http://www.prosportstransactions.com/basketball. 

```{r eval = FALSE, echo = TRUE}
count <- c(1:1291)
pageNumbers <- (count-1)*25
commonUrl <- 'http://www.prosportstransactions.com/basketball/Search/SearchResults.php?Player=&Team=&BeginDate=&EndDate=&ILChkBx=yes&Submit=Search&start='
allYearsdf <- data.frame(Date=character(),Team=character(), Acquired=character(), Relinquished=character(), Notes=character()) 
```

```{r eval = FALSE, echo = TRUE}
for (i in pageNumbers)
{
url <- paste(c(commonUrl,i),collapse='')
page <- read_html(url)
print(url)
content <- page %>%
  html_node('div.container') %>%
  html_node('table.datatable') %>%
  html_table()
headings <- head(content,1)
df<-content
colnames(df) <-c(headings)
df <- df[-1,]
x<-'•'
df$Relinquished<-sub(x,'',df$Relinquished)
df$Acquired<-sub(x,'',df$Acquired)
allYearsdf<-rbind(allYearsdf,df)
}

write.csv(allYearsdf,'basketballScrapingOutput.csv')

```

The data we have scraped needs to be cleaned. All the players who are injured are generally placed in a list known as the relinquished list. We make sure that all the data we have scraped has player names and no blank values. Apart from that, it was observed that the player names scraped has a space before the words. Hence a trim operation was done to remove the space.

```{r eval = FALSE, echo = TRUE}
newDf = read.csv('basketballScrapingOutput.csv')
newDf$Relinquished <-str_trim(newDf$Relinquished)
newDf<-newDf[!newDf$Relinquished == '',]
newDf
```

Post the retrieval,we merge the injury data we scraped with the player data from the NBA statistics package (nbaStatR). This is followed by data processing where we add the year in which the injury took place.

```{r eval = FALSE, echo = TRUE}
playerData<-dictionary_bref_players()
injuredPlayerData<-merge(newDf,playerData,by.x='Relinquished',by.y='namePlayerBREF')
date<-as.POSIXct(injuredPlayerData$Date,format = '%Y-%m-%d')
year <- as.numeric(format(date,format = '%Y'))
injuredPlayerData<-cbind(injuredPlayerData,year)
injuredPlayerData
```

For our detailed analysis, we consider the last 8 years. Beyond this range, the analysis may prove to be futile, since basketball as a game evolves fairly quickly over time.

```{r eval = FALSE, echo = TRUE}
yearsList <- sort(unique(injuredPlayerData$year),decreasing = TRUE)
yearsList <- yearsList[1:8]
```

The player data we have contains urls to the players profiles on basketball-reference.com/players. We scraped the position and hand(i.e Left Hand or Right Hand) from the urls and merged it to the injury player data we have procured.

```{r eval = FALSE, echo = TRUE}
topPlayerInjuryData <-subset(injuredPlayerData,year %in% yearsList )
injuredPlayerUrls <- topPlayerInjuryData['urlPlayerBioBREF']
positionalData<-lapply(injuredPlayerUrls[[1]], scrapPositionalData)
df<-data.frame(matrix(unlist(positionalData), nrow=length(positionalData), byrow=TRUE))
colnames(df)<-c('position1','hand')
```

The data we possess now is holistic, with injury notes over 8 years. However, on visualizing the data, it was observed that the position field contained non-atomic entries (Since a player may have played at more than one position). Hence, we made sure the position field is atomic by separating the value onto two different rows.

```{r eval = FALSE, echo = TRUE}
mergedPlayerInjuryData<-cbind(topPlayerInjuryData,df)
mergedPlayerInjuryData <- separate_rows(mergedPlayerInjuryData,position1,sep="and")
mergedPlayerInjuryData <- separate_rows(mergedPlayerInjuryData,position1,sep=",")
mergedPlayerInjuryData[!mergedPlayerInjuryData$position1 == '',]
mergedPlayerInjuryData
```

The injury notes do not give us a clear cut idea of injury the player had. Hence, we categorize the injuries of the players. The categories are obtained from the NBA injury types statistic.

```{r eval = FALSE, echo = TRUE}
injuryDataNba <- bref_injuries()
injuryClasses <- c(unique(injuryDataNba$typeInjury))
```


We create a function known as categoryDefine(), that will take in the injury notes, and the injurycategories as input parameters, and will retuirn back the most befitting category for a given entry. We then column bind this category to the injury dataset. 

```{r eval = FALSE, echo = TRUE}
notesList <- preprocessedInjuryData$Notes
injuryCategoryPlayers<-lapply(notesList, categoryDefine)
df<-data.frame(matrix(unlist(injuryCategoryPlayers), nrow=length(injuryCategoryPlayers), byrow=TRUE))
colnames(df)<-c('typeInjury')
preprocessedInjuryData<-cbind(preprocessedInjuryData,df)
```

The last step in the cleaning process was to get rid of the anomalous rows. We replaced the position Guard with Point Guard (since in a 2-2-1 zone formation a point guard is called a guard) and Forward with Small Forward (since in a 2-2-1 zone formation a point guard is called a forward). There were hardly any entries that were NA, but a few entries with '' in the column. We got rid of them by removing the rows with these values.

```{r eval = FALSE, echo = TRUE}
preprocessedInjuryData$position1[preprocessedInjuryData$position1 == 'Guard'] <- 'PointGuard'
preprocessedInjuryData$position1[preprocessedInjuryData$position1 == 'Forward'] <- 'SmallForward'
preprocessedInjuryData<-preprocessedInjuryData[!preprocessedInjuryData$Relinquished == '',]
preprocessedInjuryData<-preprocessedInjuryData[!preprocessedInjuryData$position1 == '',]

```

## d3

A small amount of tooling was required to use the NBAstatR data with d3. To make the amount of players more manageable, we 
pared down the total number of players in the dataset to just the big names on each team. There 
are a number of players that play only a few games each year and greatly skew their ranking data. So,
get the best and most notable players, we needed to look beyond ratings. First, took data from the whole season for only _starters_. 
This already eliminates a lot of players -- only reasonably good players get to start. 
Then, we took the top 6 ranks for each team by number of games started. 

```{r eval = FALSE, echo = TRUE}
starters <- teams_players_stats(seasons = 2020, types = "player", starters_bench = list("Starters")) %>% 
  unnest(dataTable) 

top_guys_bubble <- starters %>% 
  semi_join(bubble_seeding_players, by = "namePlayer") %>% #only return players with bubble data, some players opted out
  semi_join(covid_season_players_month_before, by = "namePlayer") %>% 
  group_by(slugTeam) %>% 
  slice_max(gp, n = 6)
```


Once we found the list of players we wanted, we filtered our game data from these names. 
Then, we cut down the amount of game stats to just the ones we were interested in.
Finally, we spread the data wide by time period so that data from each period would be associated with a 
single svg element. This would allow us to access this data easily, letting us read the data just once.

Being d3 beginners, we were not sure of this approach. Maybe performance would have been better to 
add/remove dots as needed rather than load them all and hide/show them? Or even split the csv into three, 
one for each time period, loading as needed. Regardless, this approach worked!
```{r eval = FALSE, echo = TRUE}
d3_data <- bubble_season_combined %>% 
  semi_join(top_guys_bubble, by = "namePlayer") %>% 
  left_join(bubble_teams, by = "slugTeam") %>% 
  select(name = "namePlayer",
         team_abb = "slugTeam",
         team_name = "nameTeam",
         ortg,
         drtg,
         netrtg,
         df_type
         ) %>% 
  pivot_wider(names_from = "df_type", values_from = c("ortg", "drtg", "netrtg"))
 
d3_data %>%
  write_csv("data/d3_data.csv")

```
